---
title: "RNAseq QC Report"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    
params:
  results: null
  min_assigned_read_count: null
---

```{r setup, include=FALSE}
library(flexdashboard)
library(ggplot2)
library(plotly)
library(DT)
library(scales)
library(GGally)
library(jsonlite)
``` 

Read Count Analysis
=====================================

Column {data-width=650}
-------------------------------------

### Read count analysis

```{r reads_plot}
results <- readRDS(paste0(params$results, "/results/rds/reads_plot.rds"))
min_assigned_read_count <- params$min_assigned_read_count

reads_plot <- ggplot(results, aes(x = sample, y = reads, fill = status)) +
  geom_hline(yintercept=c(min_assigned_read_count, 50e6)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(
    	breaks = function(x) {
      		default_breaks <- extended_breaks()(x)
      		sort(c(default_breaks, min_assigned_read_count))
    	},
    	labels = function(breaks) {
     		ifelse(breaks == min_assigned_read_count, paste0("cutoff: ", min_assigned_read_count), breaks)
    	}
  ) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 0))

ggplotly(reads_plot)
``` 


Column {data-width=350}
-------------------------------------

### Read count analysis

```{r}
datatable(results)
```


FastP Analyses
=====================================

Column {.tabset .tabset-pills}
-------------------------------------

```{r, include=FALSE}
trimmed_path <- paste0(params$results, "/results/trimmed")

files <- list.files(path = trimmed_path, pattern = "\\.json$", full.names = TRUE)

fastp_list <- lapply(files, fromJSON)

names(fastp_list) <- gsub(".json", "", basename(files))


get_quality_df <- function(x, sample_name, read, stage, status, metric) {
  df <- as.data.frame(x[[paste0("read", read, "_", stage, "_filtering")]][[metric]])
  if(nrow(df) > 0) {
    df$position <- seq_len(nrow(df))
    df$sample <- sample_name
    df$status <- status
    df
  }
}
```

### Duplication

```{r fastp_duplication}
dup_df <- tibble(
  sample = names(fastp_list),
  duplication_rate = map_dbl(fastp_list, ~ .x$duplication$rate)
)

dupp <- ggplot(dup_df, aes(x = sample, y = duplication_rate*100)) +
  geom_col(fill = "red", col="black") +
  labs(x = "Sample", y = "Rate %") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 0))
ggplotly(dupp)
```

### QC

```{r fastp_qc}
quality_df <- map2_dfr(fastp_list, names(fastp_list), \(x, n) {
  rbind(
    get_quality_df(x, n, 1, "before", "untrimmed forward", "quality_curves"),
    get_quality_df(x, n, 2, "before", "untrimmed reverse","quality_curves"),
    get_quality_df(x, n, 1, "after",  "trimmed forward","quality_curves"),
    get_quality_df(x, n, 2, "after",  "trimmed reverse","quality_curves")
  )
})


mqc <- ggplot(quality_df, aes(x = position, y = mean, color = sample)) +
  geom_line(alpha = 0.7) + ylim(0, 40) + 
  labs(x = "Position", y = "Quality score") +
  facet_wrap(~status) 

ggplotly(mqc)

```

### GC

```{r fastp_gc}
gc_df <- map2_dfr(fastp_list, names(fastp_list), \(x, n) {
  rbind(
    get_quality_df(x, n, 1, "before", "untrimmed forward", "content_curves"),
    get_quality_df(x, n, 2, "before", "untrimmed reverse","content_curves"),
    get_quality_df(x, n, 1, "after",  "trimmed forward","content_curves"),
    get_quality_df(x, n, 2, "after",  "trimmed reverse","content_curves")
  )
})

mgc <- ggplot(gc_df, aes(x = position, y = GC*100, color = sample)) +
  geom_line(alpha = 0.7) + 
  labs(x = "Position", y = "GC %") +
  facet_wrap(~status) 
ggplotly(mgc)

```

### Insert Size

```{r fastp_insert_size}
extract_insert_size <- function(x, sample_name) {
  if("insert_size" %in% names(x)) {
    hist <- x$insert_size$histogram
  
    last_nonzero <- max(which(hist != 0))
  
    hist_trimmed <- hist[seq_len(last_nonzero)]
  
    tibble(
      insert_size = seq_along(hist_trimmed),
      count = hist_trimmed,
      sample = sample_name
    )
  }
}

insert_df <- map2_dfr(fastp_list, names(fastp_list), extract_insert_size)

isp <- ggplot(insert_df, aes(insert_size, count, color = sample)) +
  geom_line(alpha = 0.7) +
  theme_minimal() +
  labs(x = "Insert size (bp)", y = "Read count")
ggplotly(isp)
```


FastQC/Falco Analyses
=====================================

Column {.tabset .tabset-pills}
-------------------------------------

```{r read falco qc, include=FALSE}
update_samples <- function(x) {
  read_type = c("untrimmed fwd", "untrimmed rev", "trimmed fwd", "trimmed rev", "untrimmed single", "trimmed single")
  names(read_type) = c("1", "2", "R1", "R2", "s", "S")
  x <- x |> extract(Sample, into = c("Sample", "read"), regex = "(.+)_(R?[sS12])$")
  x$read = factor(x$read)
  levels(x$read)=read_type[as.character(levels(x$read))]
  x
}

parse_table <- function(sec) {
  hi <- tail(grep("^#", sec$lines), 1); if (is.na(hi)) return(NULL)
  hd <- str_remove(sec$lines[hi], "^#") |> str_split("\t") |> unlist()
  dl <- sec$lines[(hi+1):length(sec$lines)]; dl <- dl[nzchar(dl)]
  if (!length(dl)) return(NULL)
  mat <- str_split_fixed(dl, "\t", length(hd))
  df  <- as.data.frame(mat, stringsAsFactors = FALSE)
  colnames(df) <- make.names(hd); df
}

all_data <- readRDS(paste0(params$results, "/results/rds/falco_qc.rds"))
```


### Per base sequence quality

```{r Per base sequence quality}
pbq_df <- map_dfr(names(all_data), function(s) {
  sec <- all_data[[s]][["Per base sequence quality"]]; if(is.null(sec)) return(NULL)
  df  <- parse_table(sec); if(is.null(df)) return(NULL)
  df$Sample <- s; df
}) |> mutate(Base=as.integer(str_extract(Base,"^[0-9]+")), Mean=as.numeric(Mean))

pbq_df <- update_samples(pbq_df)

ggplotly(ggplot(pbq_df, aes(x=Base, y=Median, group=Sample, col=Sample)) + geom_line() + facet_wrap(~read))
```


### Per-sequence quality

```{r per sequence quality}
psq_df <- map_dfr(names(all_data), function(s) {
  sec <- all_data[[s]][["Per sequence quality scores"]]; if(is.null(sec)) return(NULL)
  df  <- parse_table(sec); if(is.null(df)) return(NULL)
  df$Sample <- s; df
}) |> mutate(Quality=as.integer(Quality), Count=as.numeric(Count))

psq_df <- update_samples(psq_df)

ggplotly(ggplot(psq_df, aes(x=Quality, y=Count, col=Sample)) + geom_line() + facet_wrap(~read))
```


### GC content

```{r gc content}
gc_df <- map_dfr(names(all_data), function(s) {
  sec <- all_data[[s]][["Per sequence GC content"]]; if(is.null(sec)) return(NULL)
  df  <- parse_table(sec); if(is.null(df)) return(NULL)
  df$Sample <- s; df
}) |> mutate(GC.Content=as.integer(GC.Content), Count=as.numeric(Count))

gc_df <- update_samples(gc_df)

ggplotly(ggplot(gc_df, aes(x=GC.Content, y=Count, col=Sample)) + geom_line() + facet_wrap(~read))
```


### Duplication levels

```{r duplication level}
dup_df <- map_dfr(names(all_data), function(s) {
  sec <- all_data[[s]][["Sequence Duplication Levels"]]; if(is.null(sec)) return(NULL)
  df  <- parse_table(sec); if(is.null(df)) return(NULL)
  df$Sample <- s; df
}) |> rename_with(~str_trim(str_replace_all(.x,"\\.+"," "))) |>
  mutate(across(where(is.character) & !matches("Sample|Duplication"), as.numeric))

dup_pct_col <- grep("dedup|Percentage", colnames(dup_df), ignore.case=TRUE, value=TRUE)[1]
if (!is.na(dup_pct_col)) dup_df <- rename(dup_df, Pct_dedup = all_of(dup_pct_col))

dup_lvl_col <- grep("Duplication Level", colnames(dup_df), ignore.case=TRUE, value=TRUE)[1]
if (!is.na(dup_lvl_col)) dup_df <- rename(dup_df, DupLevel = all_of(dup_lvl_col))

dup_df <- update_samples(dup_df)

dup_df$DupLevel <- factor(dup_df$DupLevel, levels=unique(dup_df$DupLevel))
ggplotly(ggplot(dup_df, aes(x=DupLevel, y=`Percentage of total`, col=Sample, group=Sample)) + geom_line() + facet_wrap(~read))
``` 



### Adapter content

```{r adapter content}
ac_df <- map_dfr(names(all_data), function(s) {
  sec <- all_data[[s]][["Adapter Content"]]; if(is.null(sec)) return(NULL)
  df  <- parse_table(sec); if(is.null(df)) return(NULL)
  df$Sample <- s; df
}) |> pivot_longer(-c(Position,Sample), names_to="Adapter", values_to="Pct") |>
  mutate(Position=as.integer(str_extract(Position,"^[0-9]+")), Pct=as.numeric(Pct))

ac_df <- update_samples(ac_df)

ac_list <- ac_df |> group_by(Adapter) |> summarize(max = max(Pct)) |> filter(max >= 0.05) |> select(Adapter)

ggplotly(ggplot(ac_df |> filter(Adapter %in% ac_list), aes(x=Position, y=Pct, col=Sample)) + geom_line() + facet_wrap(~read+Adapter))
```







Read Assignment Analysis
=====================================

Column {data-width=350}
-------------------------------------

```{r}
dfmdet <- readRDS(paste0(params$results, "/results/rds/gene_coverage_plot.rds"))

dfm <- dfmdet$dfm
detected <- dfmdet$detected
detected2 <- dfmdet$detected2

knitr::kable(head(dfm))
```


Column {.tabset data-width=650}
-------------------------------------

### Read Assignment Comparison

```{r read_count_comp}
mat <- readRDS(paste0(params$results, "/results/rds/read_count_comp.rds"))

heatmaply::heatmaply(mat, dendrogram = "none", 
                     limits = c(0,1),
                     colors = colorRampPalette(rev(RColorBrewer::brewer.pal(11, "RdYlBu")))(100))
```


### Read Assignment Analysis

```{r gene_coverage}
gcp_colors <- c("#e0e0e0", "#faa39d", "#dbc160", "#00BA38")
names(gcp_colors) <- levels(dfm$variable)

gene_coverage_plot <- ggplot(dfm, aes(x = Replicate,
                                      y = value,
                                      fill = variable)) +
  geom_hline(yintercept = c(detected, detected2)) +
  geom_bar(stat = "identity") +
  ylab("Number of Genes") + xlab("Samples") +
  guides(fill = guide_legend(title = "Number of assigned reads")) +
  theme(axis.text.x = element_text(angle = 90, hjust = 0)) +
  facet_wrap(~Sample, scale = "free_x") +
  scale_fill_manual(values = gcp_colors)

ggplotly(gene_coverage_plot, height = 800)

```





Gene Body Coverage
=====================================

Column {data-width=650}
-------------------------------------

### RNA degradation (gene body coverage)

```{r genebody_coverage}
d <- readRDS(paste0(params$results, "/results/rds/genebody_coverage.rds"))
deg_matrix <- d$matrix
ccols = d$colors
skewclass = d$skewclass
cclass = d$cclass

deg_matrix_sorted <- deg_matrix[order(skewclass), ]
cclass_sorted <- cclass[order(skewclass), , drop = FALSE]

flat_p <- do.call(c, ccols)
for (col in names(cclass_sorted)) {
  cclass_sorted[[col]] <- paste0(col, ".", cclass_sorted[[col]])
}

heatmaply::heatmaply(
  as.matrix(deg_matrix_sorted),
  dendrogram = "none",
  show_dendrogram = c(FALSE, FALSE),
  row_side_colors = cclass_sorted,
  row_side_palette = flat_p,
  colors = colorRampPalette(rev(RColorBrewer::brewer.pal(11, "RdYlBu")))(100),
  colorbar_title = "coverage (%)"
) %>% plotly::layout(legend = list(orientation = "h", x = 0.5, xanchor = "center", y = -0.1, yanchor = "top"))
```


Size Factor QC
=====================================

Column {.tabset data-width=500}
-------------------------------------

### Size Factor QC
#### (see https://www.bioconductor.org/packages/release/bioc/vignettes/DEGreport/inst/doc/DEGreport.html)

```{r factor_plot}
df <- readRDS(paste0(params$results, "/results/rds/split_factor_plot.rds"))

factor_plot <- ggplot(df, aes(ratios, col = sample, group = replicate)) + 
    geom_vline(xintercept=0) + geom_density() + theme_bw()
ggplotly(factor_plot)

```

### by Condition

```{r split_factor_plot}
split_factor_plot <- ggplot(df, aes(ratios, group = replicate, col = sample)) + 
    geom_vline(xintercept=0) + geom_density() + theme_bw() + facet_wrap(~sample)

ggplotly(split_factor_plot)

```



Sample Correlation
=====================================

Column {data-width=1000}
-------------------------------------

### Sample Correlation Heatmap of VST counts 

```{r sample_correlation}
d <- readRDS(paste0(params$results, "/results/rds/correlation_hm.rds"))

sample_cor <- d$sample_cor
a = d$a 

col_order <- hclust(dist(sample_cor))$order

heatmaply::heatmaply(
  sample_cor,
  row_side_colors = a,
  col_side_colors = a,
  colors = colorRampPalette(rev(RColorBrewer::brewer.pal(11, "RdYlBu")))(100)
)  %>% plotly::layout(legend = list(orientation = "h", x = 0.5, xanchor = "center", y = -0.1, yanchor = "top"))
```


Sample PCA
=====================================

Column {.tabset data-width=1000}
-------------------------------------

### PC1 vs PC2

```{r PCA}
pca_res = readRDS(paste0(params$results, "/results/rds/pca_res.rds"))

# scores (samples)
scores <- as.data.frame(pca_res$rotated)
scores$sample <- rownames(scores)

# add metadata
scores <- cbind(scores, pca_res$metadata)

pca_plot <- ggplot(scores, aes(x = PC1, y = PC2, color = condition)) +
  geom_point(size = 3) +
  labs(
    x = paste0("PC1 (", round(pca_res$variance["PC1"], 1), "%)"),
    y = paste0("PC2 (", round(pca_res$variance["PC2"], 1), "%)")
  ) +
  theme_bw()

ggplotly(pca_plot)
```

### PCA Pairs plot

```{r PCA_pairs}
minpc <- min(ncol(pca_res$rotated), 5)
scores <- as.data.frame(pca_res$rotated[, 1:minpc])
scores <- cbind(scores, pca_res$metadata)

p = ggpairs(
  scores,
  columns = 1:minpc,
  aes(color = condition),
  upper = list(continuous = "points"),
  lower = list(continuous = "points"),
  diag = list(continuous = "densityDiag")
)

ggplotly(p)
```



PCA outlier
=====================================

Column {.tabset data-width=1000}
-------------------------------------

### PCA Outlier

```{r pca_outlier}
  pca_obj <- readRDS(paste0(params$results, "/results/rds/pca_outlier_2.rds"))

  pca_data <- pca_obj$pca_data
  outlier_data <- pca_obj$outlier_data

  pca_outlier <- ggplot(pca_data, aes(x = condition, y = rot)) +
        geom_point(size=0.5)  +
        geom_label(data=outlier_data, aes(label=samples), hjust = 0) +
        geom_point(data=outlier_data, col="red", size=0.5) +
        facet_wrap(~PC)
  
  ggplotly(pca_outlier)
```


### PCA Sample Distance

```{r pca_intra_sample_dist}
pca_obj <- readRDS(paste0(params$results, "/results/rds/pca_outlier.rds"))

dns_values <- pca_obj$dns_values
peaks <- pca_obj$peaks

intra_sample_dist <- ggplot(dns_values, aes(x=x, y=y)) +
    geom_vline(data = peaks, mapping = aes(xintercept = peak), col = "gray") +
    geom_line() +
    xlab("intra sample PCA distance") +
    facet_wrap(~pc)
ggplotly(intra_sample_dist)
```


